---
title: "PS5 Assignment: INFO 201"
author: "Hrudhai Umas"
date: "2023-02-13"
output: html_document
---
# PS5 Assignment RMarkdown and GGPlot

### 1) Load and Check Data

**1. (1pt) Create a new rmarkdown document with an appropriate title **

- Title: "PS5 Assignment: INFO 201"
- Author: "Hrudhai Umas"
- Date: "2023-02-13"

**2. (2pt) Load data. How many rows/columns do we have?**
```{r, message=FALSE}
library(tidyverse)

gm_data <- read.delim("gapminder.csv")

rows <- gm_data %>% 
  nrow()

cols <- gm_data %>% 
  ncol()

cat("There are", rows, "rows and", cols, "columns in this dataset!")
```
**3. (2pt) Print a small sample of the data: does it look okay? **

```{r}
peek <- gm_data %>% 
  select(1:4) %>% 
  sample_n(10)

cat("Here is a peek at our dataset! Only first 4 columns shown to make output 
more cleaner. Full dataset was checked however before!")
peek
```

### 2) Descriptive Statistics (15pt)

**1. (3pt) How many countries are there in the dataset? **
```{r}
gm_data %>% 
  filter(iso2 != " ", name != " " ) %>% 
  summarize(Unique_iso3 = n_distinct(iso3), 
            Unique_iso2 = n_distinct(iso2), 
            Unique_name = n_distinct(name))
```
**2. There are more names than iso-2 codes, and there are even more iso3**
**-codes.What is going on? Can you find it out?**

- **(5pt) Find how many names are there for each iso-2 code. Are there any** 
**iso-2 codes that correspond to more than one name? What are these countries?**
```{r}
count_of_occur <- gm_data %>% 
  group_by(iso2) %>% 
  summarize(number_of_occurrences = n_distinct(name)) %>% 
  arrange(desc(number_of_occurrences))

value <- count_of_occur %>% 
  filter(number_of_occurrences > 1) %>%
  pull(iso2)

gm_data %>% 
  group_by(iso2) %>% 
  filter(iso2 == value) %>% 
  filter(name != "") %>% 
  pull(name) %>% 
  head(1)
```

- What we see happened here was that **Namibia's iso-2 code is NA**, which RStudios
thought meant **NullValue** and that is why it got duplicated out with another
country which had no name and no iso-2 code.

- **(5pt) Now repeat the same for the name and iso3-code. Are there country**
**names have more than one iso3-code? What are these countries?**
```{r}
gm_data %>% 
  group_by(iso3) %>% 
  summarize(number_of_occurrences = n_distinct(name)) %>% 
  arrange(desc(number_of_occurrences))

##find iso-3 that does not have a name
gm_data %>% 
  group_by(iso3) %>% 
  filter(name == "") %>% 
  summarise(country = unique(iso3) ) %>% 
  select(country)
```

- The 4 countries as seen above all have an iso-3 name but no country name!
Hence, that is why there exists more iso-3 than country names!

**3. (2pt) What is the minimum and maximum year in these data?**

```{r}
min_year <- gm_data %>% 
  summarize(min_year = min(time, na.rm = TRUE)) %>% 
  pull(min_year)

max_year <- gm_data %>% 
  summarize(max_year = max(time, na.rm = TRUE)) %>% 
  pull(max_year)

cat("The minimum year in these data is", min_year, "and the maximum year is",
    max_year, ".")

```

### 2) CO2 Emissions (30pt)

**1. (2pt) How many missing co2 emissions are there for each year?** 
**Analyze both missing CO2 and co2_PC. Which years have most missing data?**
```{r}
co2_missing <- gm_data %>% 
  group_by(time) %>%
  summarize(missing_co2 = sum(is.na(co2)), missing_co2_PC = 
              sum(is.na(co2_PC))) %>% 
  arrange(desc(missing_co2), desc(time))

drop_na(co2_missing) 
```
- As seen in the data, the years _2017-2019_ have the most amount of missing 
data with **217.** Following are the years _1961 and 1960_(missing **60 each**). 

**2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China,** 
**and India. Add a few more countries of your choice. Explain what do you see.**
```{r, message = FALSE, warning= FALSE}
gm_data %>%
  filter(name == "United States of America" |name == "China" | name == "India" |
        name == "Brazil" | name == "Italy") %>%
  select(time, co2, name) %>%
  ggplot(aes(x = time, y = co2, color = name)) +
  geom_line() +
  labs(x = "Year", y = "CO2 Emissions (kt)", color = "Country")
```

- Based on the graph, we can see that **China has seen an increase** in C02 
emissions since the year **1960** and has **emitted the most with close to 
**1.0e+07kt** but is seen to have **decreased emissions in the past years**. 
The **United States** is next, with an overall trend showing **an increase** 
but have also seen a **decrease of emissions** in the past years. **India** 
can be seen to have an overall **continuous increase in emissions** with 
_no decreases_. The countries of **Italy and Brazil** are all in the lower end 
of the emissions spectrum and have a **steady increase** in emissions for the 
most part. 

**3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC)** 
**Make a similar plot of the same countries. What does this figure suggest?**
```{r, message = FALSE, warning = FALSE}
gm_data %>%
  filter(name == "United States of America" |name == "China" | name == "India" 
        | name == "Brazil" | name == "Italy") %>%
  select(time, co2_PC, name) %>%
  ggplot(aes(x = time, y = co2_PC, color = name)) +
  geom_line() +
  labs(x = "Year", y = "CO2 Emissions (metric tons per capita)", color = 
         "Country")
```

- This graphs suggest that the **United States has the highest carbon emissions** 
per capita (metric tons) with close to _17 metric tons_ but is steadily 
_decreasing_. After that we see **China have the second most per**
**capita emissions** with a _steady increase_ but has seen a _decrease in the_ 
_past years_. After, we see **Italy have been decreasing** in the last couple of 
years with the amount of emissions. We then see Brazil and lastly **India which** 
**is seen to have a steady increase**, but, still the lowest with less than 
_5 metric tons carbon emissions per capita_. 

**4. (6pt) Compute average CO2 emissions per capita across the continents** 
**(assume region is the same as continent). Comment what do you see.**
```{r}
gm_data %>% 
  filter(time == 1960 | time == 2016, !is.na(co2_PC), !is.na(region) ) %>%
  filter(region != "") %>% 
  group_by(time, region) %>% 
  summarize(average_CO2_PC = mean(co2_PC, na.rm = TRUE) ) 
```
```{r}
##In this table, we can see: 
##Africa: Increase in average CO2 Per capita from 1960 to 2016
##Americas: Decrease in average CO2 Per capita from 1960 to 2016.
#Asia: Increase in average CO2 Per capita from 1960 to 2016
##Europe: Increase in average CO2 Per capita from 1960 to 2016.
##Oceania: Increase in average CO2 Per capita from 1960 to 2016.

##These increases can be associated with the heavy use of fossil fuels in 
##the new modern advancements in technology that we have today like cars. The
##regions that have seen a decrease in per capita CO2 can be seen to be using 
##less fossil fuels or more efficient technology to reduce emissions. 
```


5. **(7pt) Make a bar plot where you show the previous results–average CO2** 
**emissions per capita across continents in 1960 and 2016**
```{r, message= FALSE}
gm_data %>% 
  filter(time == 1960 | time == 2016, !is.na(co2_PC), !is.na(region) ) %>%
  filter(region != "") %>% 
  group_by(time, region) %>% 
  summarize(average_CO2_PC = mean(co2_PC, na.rm = TRUE) ) %>% 
  ggplot(aes(x = region, y = average_CO2_PC, fill = as.factor(time))) + 
  geom_col(position =  "dodge") + 
  labs(title = "Average CO2 Emissions per Capita by Year and Region(Continent)",
       x = "Region(Continent)", y = "Average CO2 Emissions Per Capita
       (metric tons per capita)") +
  scale_fill_discrete(name = "Year")
```

- Overall, this graph can lead us to conclude that **all regions** have seen an 
increase in the emissions of CO2 per capita **except for the Americas**. Reasons 
could be because of the **lower usage of fossil fuels** 
or **advancements in technology**to lead to the decrease in the Americas region.
The **increase can be explained** due to the fact that many countries rely on 
**heavy usage of fossil fuels as energy** and other activities like 
**transportation.**

**6. Which countries are the three largest, and three smallest CO2 emitters** 
**(in terms of CO2 per capita) in 2016 for each continent?**
```{r, message = FALSE}
gm_data %>%
  filter(time == 2016, !is.na(co2_PC), !is.na(region)) %>%
  filter(name != "") %>% 
  group_by(region, name) %>%
  summarize(co2_pc_data = co2_PC) %>%
  arrange(region, desc(co2_pc_data)) %>%
  mutate(rank = row_number()) %>%
  filter(rank <= 3 | rank >= n() - 2) %>%
  ungroup() %>%
  arrange(region, rank)
```

- Using 2016 data due to the fact 2019 is all N/A. Here are the countries with 
the three largest and three smallest CO2 emitters:

- Africa(Largest): South Africa, Libya, Seychelles
- Africa(Smallest): Burundi, Somalia, Congo: Democratic Republic of the

- Americas(Largest): Trinidad and Tobago, United States of America, Canada
- Americas(Smallest): Honduras, Nicaragua, Haiti

- Asia(Largest): Qatar, Kuwait, Bahrain
- Asia(Smallest): Yemen, Nepal, Afghanistan

- Europe(Largest): Gibraltar, Luxembourg, Faroe Island
- Europe(Smallest): Moldova: Republic of, Albania, Liechtenstein 

- Oceania(Largest): New Caledonia, Australia, Palau,   
- Oceania(Smallest): Kiribati, Vanuatu, Solomon Islands 

### 4) GDP Per Capita (50pt)

**1. (8pt) Make a scatterplot of GDP per capita versus life expectancy**
**by country, using data for 1960.**
```{r}
gm_data %>%
  filter(time == 1960, !is.na(GDP_PC), !is.na(lifeExpectancy), name != "") %>%
  ggplot(aes(x = GDP_PC, y = lifeExpectancy, color = region, 
             size = totalPopulation)) +
  geom_point() +
  scale_size(range = c(1, 5)) +
  labs(title = "Life Expectancy(years) and GDP Per Capita(USD) in 1960", 
       x = "GDP Per Capita (USD)", y = "Life Expectancy (years)", 
       color = "Region", size = "Population Size")
```

- What I see is that many countries from the **European regions have higher GDPs**
and this can also be associated to the fact that they **have higher life expectancy**.
However this not always the case as many in the **Americas have decent(55+) years**
of life expectancy. However many regions such as **Africa and Asia** have countries
whose **GDP is less than 5000** and have a life expectancy _less_ than **50 years** as
it may be caused by the **lack of development in that region**.

**2. (4pt) Make a similar plot, but this time use 2019 data only.**
```{r}
gm_data %>%
  filter(time == 2019, !is.na(GDP_PC), !is.na(lifeExpectancy), name != "") %>%
  ggplot(aes(x = GDP_PC, y = lifeExpectancy, color = region, 
             size = totalPopulation)) +
  geom_point() +
  scale_size(range = c(1, 5)) +
  labs(title = "Life Expectancy(years) and GDP Per Capita(USD) in 2019",
       x = "GDP Per Capita (USD)", y = "Life Expectancy (years)", 
       color = "Region", size = "Population Size")
```

- In the **2019 graph**, we can see that **life expectancy is above 50** for *all countries*
which can be explained due to the **advancement in technology and medical** 
**advancements** that helps people _live longer_. The **highest life expectancy** is 
still many **European nations as they have more GDP**. Higher GDP Per Capita
means it **allows for a better standard of living** and life expectancy! After Europe, it 
can be found that the **Americas** have the next **higher life expectancy** with 
also a decently high GDP per capita. **Africa and Oceania still is in the lower**
end of the GDP per capita and still has a **lower life expectancy** than other 
regions for the most part.

**3. (6pt) Compare these two plots and comment what do you see. How has world** 
**developed through the last 60 years?**

-  There exist a **positive relationship between GDP Per Capita and**
**life expectancy** _for both plots_. However, we see that the **2019 plot** 
shows a **higher life expectancy and GDP Per Capita** compared to the _1960 plot_. 
We can say that **all countries** have made progress in terms of many fields.
For example, there may have been a **improvement in healthcare** such that there 
is **better medical technology** and **improved public health measures** that 
leads to _fewer deaths from preventable diseases_. Also a growing economy can be 
seen from **better investments in healthcare**, **education**, and other services 
to boost **standard of living**. More **advance technology can make it easier** to 
**diagnose diseases** and create _affordable and effective pharmaceuticals_. 
These are just some ways the world has developed through the last 60 years. 

**4. (6pt) Compute the average life expectancy for each continent in 1960 and** 
**2019. Do the results fit with what do you see on the figures?**
```{r, message = FALSE}
gm_data %>% 
  filter(time == 1960 | time == 2019, !is.na(lifeExpectancy), name != "") %>% 
  group_by(time, region) %>% 
  summarize(avg_le = mean(lifeExpectancy, na.rm = TRUE)) %>% 
  arrange(time, desc(avg_le))
```
```{r}
##1960: Europe: 68.28254: This value makes sense as many(all) countries for 
##European regions are above 60 years for life expectancy for 1960!

##1960: Americas: 58.64651:This value makes sense as many countries of this 
##Americas region are between 40 and 65 years. While there are values above
##65 years, there are not enough to pull the average up in 1960.

##1960: Oceania: 56.39613: This value makes sense as many countries of this 
##Oceania region are between 35 and 60 years. There are countries above 60 years, 
##but not enough to pull the average up in 1960!

##1960: Asia: 51.64931: This value makes sense as many countries of this 
##Asian region are between 40 and 55 years. There are countries above 55 years, 
##but not enough to pull the average up in 1960.

##1960: Africa: 41.46600: This value makes sense as many countries of this 
##African region are between 30 and 50 years. There seems to be only one other
##country that is above 50 years of life expectancy. 


##2019: Europe: 79.35714: This value makes sense as many countries of this 
##region have a life expectancy between 70 and 85 years, hence why the average
##is high.

##2019: Americas: 75.83206: This value makes sense as many countries of this 
##region has a life expectancy between 65 and 85, hence why the Americas also 
##have a high average life expectancy. 

##2019: Asia: 74.61739: This value makes sense as many countries of this region 
##have a life expectancy between 65 and 80 years. It makes sense that this is 
##the average as most countries life expectancy is around the 70 years mark!

##2019: Oceania: 73.52827: This value makes sense as many countries of this
##region have a life expectancy of 65 to 75 years. There are 2 countries
##which have a life expectancy above 80 and near 85, which may push the 
##average up to 73!

##2019: Africa: 64.11014: This value makes sense as many countries of this region
##have a life expectancy of 55 to 70 with some countries having more than 70 years.

##Overall, the averages for both make sense as there has to have been an increase
##in life expectancy from 1960 and 2019 with all the advancemtns in those 60 years. 
```


**5. (8pt) Compute the average LE growth from 1960-2019 across the continents.** 
**Show the results in the order of growth. Explain what do you see.**
```{r, warning= FALSE, message=FALSE}
gm_data %>%
  filter(!is.na(lifeExpectancy), time >= 1960 & time <= 2019) %>%
  filter(name != "") %>%
  group_by(region, name) %>%
  summarize(lifeExpDiff = lifeExpectancy - lag(lifeExpectancy)) %>%
  group_by(region) %>%
  summarize(avgLEgrowth = mean(lifeExpDiff, na.rm = TRUE)) %>%
  arrange(avgLEgrowth)
```

**6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019.** 
**Try to put both histograms on the same graph, see how well you can do it!)**
```{r}
gm_data %>% 
  filter(time == 1960 | time == 2019, !is.na(GDP_PC)) %>% 
  ggplot(aes(x = GDP_PC, fill = as.factor(time))) + 
  geom_histogram(alpha = .5, position = "stack", bins = 25) +
  labs(title = "Histogram of GDP per Capita in 1960 and 2019",
       x = "GDP per Capita(USD)", y = "Frequency",
       fill = "Year") +
  scale_fill_manual(values = c("blue", "red"))
```

- This histogram shows the most important information of how **wealth has increased**
in the 60 years. For example, there exist only data from 2019 for GDP for 
**capitas over 35k USD**. While there are not a lot of frequency for these values, 
there still exist **no frequency of GDP greater than 35k for 1960**. Overall, there
are still high frequencies for GDP from near _0 GDP per capita USD_ to
_30k GDP per capita USD_!

**7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and** 
**in 2019? (When counting from top.)**
```{r}
# Rank US in terms of life expectancy in 1960
gm_data %>% 
  filter(time == 1960, !is.na(lifeExpectancy)) %>%
  filter(name != "") %>% 
  arrange(desc(lifeExpectancy)) %>% 
  mutate(ranking = rank(desc(lifeExpectancy))) %>% 
  filter(name == "United States of America") %>% 
  summarise(name, ranking)
  
# Rank US in terms of life expectancy in 2019
gm_data %>% 
  filter(time == 2019, !is.na(lifeExpectancy)) %>% 
  arrange(desc(lifeExpectancy)) %>% 
  mutate(ranking = rank(desc(lifeExpectancy))) %>% 
  filter(name == "United States of America") %>% 
  summarise(name, ranking)

```

**8. (6pt) If you did this correctly, you noticed that US ranking has been** 
**falling quite a bit. But we also have more countries in 2019–what about the** 
**relative rank divided by the corresponding number of countries that have LE** 
**data in the corresponding year?**
```{r}
# Find the number of countries that have LE data for 1960
number_of_countries_that_have_LE_data_1960 <- gm_data %>% 
  filter(time == 1960 & !is.na(lifeExpectancy)) %>% 
  filter(name != "") %>% 
  summarise(number_of_countries = n())

# Find the rank of the US in 1960 and divide by the number of countries
us_relative_rank_1960 <- gm_data %>% 
  filter(time == 1960) %>% 
  filter(name != "") %>% 
  mutate(rank = rank(desc(lifeExpectancy))) %>% 
  filter(name == "United States of America") %>% 
  pull(rank) / number_of_countries_that_have_LE_data_1960$number_of_countries

number_of_countries_that_have_LE_data_2016 <- gm_data %>% 
  filter(time == 2019 & !is.na(lifeExpectancy)) %>%
  filter(name != "") %>% 
  summarise(number_of_countries = n())

us_relative_rank_2019 <- gm_data %>% 
  filter(time == 2019) %>%
  filter(name != "") %>% 
  mutate(rank = rank(desc(lifeExpectancy))) %>% 
  filter(name == "United States of America") %>% 
  pull(rank) / number_of_countries_that_have_LE_data_2016$number_of_countries

cat("Relative rank of US in 1960:", us_relative_rank_1960, "\n")
cat("Relative rank of US in 2019:", us_relative_rank_2019, "\n")
```



## Reflection Questions

**1. Finally tell us how many hours did you spend on this PS.**

- I spent a total 11 hours on this PS!
